---
title: "Revision canonical simulation data generation"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
options(stringsAsFactors = FALSE)
```

# Purpose

This Rmd starts from the original TedSim-based `data_generation.Rmd` logic and converts it into a batch data-generation pipeline for the Genome Research revision.

This version generates **only the canonical synthetic datasets corresponding to the results already reported in the original manuscript and Supplementary Material**:

- main Fig. 3 mutation-rate benchmark;
- main Fig. 3 stochastic-dropout benchmark;
- Supplemental Fig. S3 target-site-number benchmark;
- Supplemental Fig. S3 sample-size benchmark;
- Supplemental Fig. S4 no-dropout benchmark.

It does **not** generate new reviewer-response experiments such as LAML, TiDeTree/TideTree, ablation, runtime, or hyperparameter-sensitivity datasets. Those should be added later as separate pools, after the original-result datasets are regenerated and verified.

It also does **not** generate the C. elegans semi-synthetic benchmark, because that analysis requires the real C. elegans expression matrix and the known C. elegans lineage tree rather than the generic TedSim state-tree generator used here.

The output remains method-agnostic: every simulated case is saved once under `data/input/simulated/<pool>/<case_id>/`, and downstream converters should translate the canonical files into algorithm-specific formats for FateScape, LinRace, LinTIMaT, NJ, Cassiopeia, LAML, or other methods.

# User settings

```{r user-settings}
# =========================
# Main switches
# =========================
library(FateScape)
RUN_GENERATION <- TRUE

# Use "1:2" for smoke test; use "all" after inspection.
CASE_INDICES <- "all"


OUT_ROOT <- "D:/input"

# Default rerun behavior:
# - SKIP_EXISTING_CASES = TRUE prevents reruns from touching completed case folders.
# - Set SKIP_EXISTING_CASES = FALSE and OVERWRITE_CASES = TRUE only when you intentionally want regeneration.
SKIP_EXISTING_CASES <- TRUE
OVERWRITE_CASES <- FALSE
STOP_ON_ERROR <- FALSE

# =========================
# Local package paths
# =========================

# Your modified TedSim path.
TEDSIM_LOCAL_PATH <- "D:/Downloads/TedSim"

# Optional FateScape package path, used only if you want to call its imputation function.
# If unavailable, the script will still generate barcode_observed and use it as barcode_imputed_fatescape.
FATESCAPE_LOCAL_PATH <- NULL
# FATESCAPE_LOCAL_PATH <- "D:/projects/FateScape/FateScape"

# =========================
# Simulation grid switches
# =========================
# These switches correspond only to the original manuscript figures.
# New revision experiments, such as LAML, TiDeTree/TideTree, ablation, runtime,
# and hyperparameter sensitivity, should be generated in separate scripts/pools.

INCLUDE_MAIN_FIG3_MUTATION <- TRUE
INCLUDE_MAIN_FIG3_DROPOUT <- TRUE
INCLUDE_FIGS3_TARGET_SITES <- TRUE
INCLUDE_FIGS3_SAMPLE_SIZE <- TRUE
INCLUDE_FIGS4_NO_DROPOUT <- TRUE

# =========================
# Canonical defaults inferred from the original manuscript/SI and this generator
# =========================
# The original manuscript/SI explicitly reports:
# - mutation-rate levels: 0.1, 0.2, 0.3
# - stochastic-dropout levels: 0.1, 0.2, 0.3
# - sample-size levels: 512, 1024, 2048
# - target-site levels: 8, 12, 16
# - no-dropout mutation-rate levels: 0.05, 0.1, 0.2
# - heritable dropout fixed at 0.05, except for the no-dropout benchmark
# - asymmetric division probability fixed at 0.8
#
# Non-varying default values used to make one-factor-at-a-time datasets:
DEFAULT_NCELLS <- 1024
DEFAULT_N_CHAR <- 12
DEFAULT_MU <- 0.2
DEFAULT_P_DS <- 0.2
DEFAULT_P_D <- 0.05
NO_DROPOUT_P_DS <- 0
NO_DROPOUT_P_D <- 0

# The manuscript says "independent simulation replicates" but does not state the number.
# The batch-generation code uses 10 replicates by default; change this only if you
# confirm that the original plotting data used another replicate count.
N_REPLICATES <- 3

# If TRUE, use True2ObservedCounts when available. Otherwise save true counts as expression.
SIMULATE_OBSERVED_COUNTS <- FALSE

# Number of retries for TedSim::SimulateCIFs when a random state walk fails.
SIMULATE_CIFS_RETRIES <- 3
MAX_WALK_CANDIDATES <- c(3)

GLOBAL_SEED <- 319
```

# Load packages

```{r load-packages}
if (!requireNamespace("ape", quietly = TRUE)) stop("Please install ape.")
library(ape)

if (!requireNamespace("devtools", quietly = TRUE)) {
  stop("Please install devtools first: install.packages('devtools')")
}

if (!is.null(TEDSIM_LOCAL_PATH) && dir.exists(TEDSIM_LOCAL_PATH)) {
  message("Loading local TedSim from: ", TEDSIM_LOCAL_PATH)
  devtools::load_all(TEDSIM_LOCAL_PATH, quiet = TRUE)
} else {
  message("Loading installed TedSim package.")
  library(TedSim)
}

if (!is.null(FATESCAPE_LOCAL_PATH) && dir.exists(FATESCAPE_LOCAL_PATH)) {
  message("Loading local FateScape from: ", FATESCAPE_LOCAL_PATH)
  devtools::load_all(FATESCAPE_LOCAL_PATH, quiet = TRUE)
}

message("R working directory: ", normalizePath(getwd(), winslash = "/", mustWork = FALSE))
OUT_ROOT <- normalizePath(OUT_ROOT, winslash = "/", mustWork = FALSE)
message("Absolute OUT_ROOT: ", OUT_ROOT)
dir.create(OUT_ROOT, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(OUT_ROOT, "_manifests"), recursive = TRUE, showWarnings = FALSE)
```

# Helper functions

```{r helper-functions}
p_tag <- function(x) {
  gsub("\\.", "p", as.character(x))
}

parse_case_indices <- function(case_indices, n_cases) {
  if (identical(case_indices, "all")) {
    return(seq_len(n_cases))
  }
  if (grepl(":", case_indices)) {
    parts <- strsplit(case_indices, ":", fixed = TRUE)[[1]]
    return(seq.int(as.integer(parts[1]), as.integer(parts[2])))
  }
  as.integer(strsplit(case_indices, ",")[[1]])
}

rbind_fill_base <- function(x, y) {
  # Base-R equivalent of dplyr::bind_rows for robust batch manifests.
  if (is.null(x) || nrow(x) == 0) return(y)
  if (is.null(y) || nrow(y) == 0) return(x)

  all_cols <- union(names(x), names(y))
  for (nm in setdiff(all_cols, names(x))) x[[nm]] <- NA
  for (nm in setdiff(all_cols, names(y))) y[[nm]] <- NA

  rbind(x[, all_cols, drop = FALSE], y[, all_cols, drop = FALSE])
}

safe_write_csv <- function(x, file, row.names = FALSE) {
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  message("Writing CSV: ", normalizePath(file, winslash = "/", mustWork = FALSE))
  utils::write.csv(x, file = file, row.names = row.names, quote = FALSE)
  if (!file.exists(file)) stop("Failed to write file: ", file)
  invisible(file)
}

safe_write_tsv <- function(x, file, row.names = FALSE) {
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  message("Writing TSV: ", normalizePath(file, winslash = "/", mustWork = FALSE))
  utils::write.table(x, file = file, sep = "\t", row.names = row.names,
                     col.names = TRUE, quote = FALSE, na = "NA")
  if (!file.exists(file)) stop("Failed to write file: ", file)
  invisible(file)
}

safe_write_tree <- function(tree, file) {
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  message("Writing tree: ", normalizePath(file, winslash = "/", mustWork = FALSE))
  ape::write.tree(tree, file = file)
  if (!file.exists(file)) stop("Failed to write tree: ", file)
  invisible(file)
}

safe_save_rds <- function(x, file) {
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  message("Writing RDS: ", normalizePath(file, winslash = "/", mustWork = FALSE))
  saveRDS(x, file = file)
  if (!file.exists(file)) stop("Failed to write RDS: ", file)
  invisible(file)
}

standardize_barcode_matrix <- function(x, cell_ids = NULL, prefix = "S") {
  x <- as.data.frame(x, check.names = FALSE)
  x[] <- lapply(x, function(z) {
    suppressWarnings(as.integer(as.character(z)))
  })
  x <- as.matrix(x)
  storage.mode(x) <- "integer"
  colnames(x) <- paste0(prefix, seq_len(ncol(x)))
  if (!is.null(cell_ids)) rownames(x) <- cell_ids
  x
}

is_missing_barcode <- function(x) {
  is.na(x) | x < 0
}

add_controlled_stochastic_dropout <- function(barcode_true, p_ds, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  bc <- barcode_true
  dropout_type <- matrix("none", nrow = nrow(bc), ncol = ncol(bc),
                         dimnames = dimnames(bc))

  heritable <- is_missing_barcode(bc)
  dropout_type[heritable] <- "heritable"

  stochastic_candidates <- !heritable
  stochastic_mask <- matrix(stats::runif(length(bc)) < p_ds,
                            nrow = nrow(bc), ncol = ncol(bc),
                            dimnames = dimnames(bc)) & stochastic_candidates

  bc[stochastic_mask] <- NA_integer_
  dropout_type[stochastic_mask] <- "stochastic"

  list(barcode_observed = bc, dropout_type = dropout_type)
}

barcode_qc <- function(barcode_true, params = NULL) {
  n_mut <- sum(!is.na(barcode_true) & barcode_true > 0)
  n_wt <- sum(!is.na(barcode_true) & barcode_true == 0)
  n_missing <- sum(is.na(barcode_true))
  n_total <- length(barcode_true)

  mutation_fraction <- n_mut / n_total
  missing_fraction <- n_missing / n_total

  message(
    "Barcode QC: mutated=", n_mut,
    ", wildtype=", n_wt,
    ", missing=", n_missing,
    ", mutation_fraction=", signif(mutation_fraction, 4),
    ", missing_fraction=", signif(missing_fraction, 4)
  )

  if (n_mut == 0) {
    if (!is.null(params)) print(params)
    stop("barcode_true contains no mutated entries. This case is invalid.")
  }

  list(
    n_mutated_entries_true = n_mut,
    n_wildtype_entries_true = n_wt,
    n_missing_entries_true = n_missing,
    mutation_fraction_true = mutation_fraction,
    missing_fraction_true = missing_fraction
  )
}

make_state_lineages <- function(cell_state_tree) {
  n_tips <- length(cell_state_tree$tip.label)
  edge <- cell_state_tree$edge
  root <- setdiff(edge[, 1], edge[, 2])
  if (length(root) != 1) stop("State tree must have a unique root.")
  root <- root[1]

  parent <- setNames(edge[, 1], edge[, 2])

  lineages <- list()
  for (tip in seq_len(n_tips)) {
    path <- tip
    cur <- tip
    while (as.character(cur) %in% names(parent)) {
      cur <- parent[as.character(cur)]
      path <- c(cur, path)
      if (cur == root) break
    }
    lineages[[paste0("L", tip)]] <- as.integer(path)
  }
  lineages
}

state_depths <- function(cell_state_tree) {
  edge <- cell_state_tree$edge
  root <- setdiff(edge[, 1], edge[, 2])
  if (length(root) != 1) stop("State tree must have a unique root.")
  root <- root[1]
  all_nodes <- seq_len(length(cell_state_tree$tip.label) + cell_state_tree$Nnode)
  depth <- rep(NA_integer_, length(all_nodes))
  names(depth) <- all_nodes
  depth[as.character(root)] <- 0L

  changed <- TRUE
  while (changed) {
    changed <- FALSE
    for (i in seq_len(nrow(edge))) {
      p <- edge[i, 1]; c <- edge[i, 2]
      if (!is.na(depth[as.character(p)]) && is.na(depth[as.character(c)])) {
        depth[as.character(c)] <- depth[as.character(p)] + 1L
        changed <- TRUE
      }
    }
  }
  depth
}

state_children_map <- function(cell_state_tree) {
  edge <- cell_state_tree$edge
  split(edge[, 2], edge[, 1])
}

is_descendant_or_equal <- function(cell_state_tree, ancestor, node) {
  if (ancestor == node) return(TRUE)
  edge <- cell_state_tree$edge
  parent <- setNames(edge[, 1], edge[, 2])
  cur <- node
  while (as.character(cur) %in% names(parent)) {
    cur <- parent[as.character(cur)]
    if (cur == ancestor) return(TRUE)
  }
  FALSE
}

state_step_distance <- function(cell_state_tree, ancestor, node) {
  if (!is_descendant_or_equal(cell_state_tree, ancestor, node)) return(NA_integer_)
  dep <- state_depths(cell_state_tree)
  as.integer(dep[as.character(node)] - dep[as.character(ancestor)])
}

simulate_cifs_retry <- function(
  ncells, cell_state_tree, p_a, n_CIF, n_diff, step,
  p_d, Sigma, mu, N_char, max_walk_candidates, SIF_res,
  unif_on = FALSE, n_retries = 3, seed = 1
) {
  last_error <- NULL

  for (attempt in seq_len(n_retries)) {
    for (mw in max_walk_candidates) {
      set.seed(seed + attempt - 1L)
      message("Trying SimulateCIFs: attempt=", attempt,
              ", seed=", seed + attempt - 1L,
              ", max_walk=", mw)
      ans <- tryCatch(
        {
          SimulateCIFs(
            ncells,
            cell_state_tree,
            p_a = p_a,
            n_CIF = n_CIF,
            n_diff = n_diff,
            step = step,
            p_d = p_d,
            Sigma = Sigma,
            mu = mu,
            N_char = N_char,
            max_walk = mw,
            SIF_res = SIF_res,
            unif_on = unif_on
          )
        },
        error = function(e) {
          last_error <<- conditionMessage(e)
          message("  SimulateCIFs failed: ", last_error)
          NULL
        }
      )
      if (!is.null(ans)) {
        attr(ans, "simulate_attempt") <- attempt
        attr(ans, "max_walk_used") <- mw
        attr(ans, "seed_used") <- seed + attempt - 1L
        return(ans)
      }
    }
  }
  stop("SimulateCIFs failed after retries. Last error: ", last_error)
}

#try_fatescape_imputation <- function(barcode_observed, N_char, ncells, r_n) {
#  if (!exists("imputation_dropout_alter", mode = "function", inherits = TRUE)) {
#    if (exists("dropout_imputation_alter", mode = "function", inherits = TRUE)) {
#      f <- get("dropout_imputation_alter", mode = "function", inherits = TRUE)
#    } else {
#      warning("No FateScape imputation function found. Returning observed barcode as barcode_imputed.")
#     return(barcode_observed)
#   }
#  } else {
#    f <- get("imputation_dropout_alter", mode = "function", inherits = TRUE)
#  }

#  ans <- tryCatch(
#    f(barcode_observed, N_char, ncells, r_n),
#    error = function(e) {
#      warning("FateScape imputation failed: ", conditionMessage(e),
#              ". Returning observed barcode as barcode_imputed.")
#      barcode_observed
#    }
#  )
#  ans <- as.matrix(ans)
#  rownames(ans) <- rownames(barcode_observed)
#  colnames(ans) <- colnames(barcode_observed)
#  ans
#}

make_character_matrix_for_external <- function(barcode, missing_code = -1L) {
  x <- barcode
  x[is.na(x)] <- missing_code
  storage.mode(x) <- "integer"
  as.data.frame(x, check.names = FALSE)
}
```

# Build parameter grid

```{r build-grid}
make_case_id <- function(pool, ncells, N_char, mu, p_ds, p_d, replicate) {
  paste0(
    pool,
    "_n", sprintf("%04d", ncells),
    "_char", N_char,
    "_mu", p_tag(mu),
    "_sd", p_tag(p_ds),
    "_hd", p_tag(p_d),
    "_rep", sprintf("%03d", replicate)
  )
}

base_grid <- function(pool, figure, panel, x_variable,
                      ncells, N_char, mu, p_ds, p_d, replicates) {
  g <- expand.grid(
    pool = pool,
    figure = figure,
    panel = panel,
    x_variable = x_variable,
    ncells = ncells,
    N_char = N_char,
    mu = mu,
    p_ds = p_ds,
    p_d = p_d,
    replicate = seq_len(replicates),
    stringsAsFactors = FALSE
  )
  g
}

grid_list <- list()

# Main Fig. 3A-C: mutation-rate benchmark.
# Variable: barcode mutation rate mu = 0.1, 0.2, 0.3.
# Non-varying defaults: n = 1024 cells, 16 target sites, stochastic dropout = 0.2,
# heritable dropout = 0.05.
if (INCLUDE_MAIN_FIG3_MUTATION) {
  grid_list$fig3_mutation <- base_grid(
    pool = "fig3_mutation",
    figure = "Fig3",
    panel = "A-C",
    x_variable = "mutation_rate",
    ncells = DEFAULT_NCELLS,
    N_char = DEFAULT_N_CHAR,
    mu = c(0.1, 0.2, 0.3),
    p_ds = DEFAULT_P_DS,
    p_d = DEFAULT_P_D,
    replicates = N_REPLICATES
  )
}

# Main Fig. 3D-F: stochastic-dropout benchmark.
# Variable: stochastic dropout p_ds = 0.1, 0.2, 0.3.
# Non-varying defaults: n = 1024 cells, 16 target sites, mutation rate = 0.2,
# heritable dropout = 0.05.
if (INCLUDE_MAIN_FIG3_DROPOUT) {
  grid_list$fig3_dropout <- base_grid(
    pool = "fig3_dropout",
    figure = "Fig3",
    panel = "D-F",
    x_variable = "stochastic_dropout_rate",
    ncells = DEFAULT_NCELLS,
    N_char = DEFAULT_N_CHAR,
    mu = DEFAULT_MU,
    p_ds = c(0.1, 0.2, 0.3),
    p_d = DEFAULT_P_D,
    replicates = N_REPLICATES
  )
}

# Supplemental Fig. S3A-C: target-site-number benchmark.
# Caption states that these datasets use 1,024 observed cells and varying target sites.
# Variable: N_char = 8, 12, 16.
if (INCLUDE_FIGS3_TARGET_SITES) {
  grid_list$figs3_target_sites <- base_grid(
    pool = "figs3_target_sites",
    figure = "FigS3",
    panel = "A-C",
    x_variable = "number_of_target_sites",
    ncells = 1024,
    N_char = c(8, 12, 16),
    mu = DEFAULT_MU,
    p_ds = DEFAULT_P_DS,
    p_d = DEFAULT_P_D,
    replicates = N_REPLICATES
  )
}

# Supplemental Fig. S3D-F: sample-size benchmark.
# Caption states that these datasets use 16 target sites and varying observed cells.
# Variable: ncells = 512, 1024, 2048.
if (INCLUDE_FIGS3_SAMPLE_SIZE) {
  grid_list$figs3_sample_size <- base_grid(
    pool = "figs3_sample_size",
    figure = "FigS3",
    panel = "D-F",
    x_variable = "number_of_cells",
    ncells = c(512, 1024, 2048),
    N_char = 16,
    mu = DEFAULT_MU,
    p_ds = DEFAULT_P_DS,
    p_d = DEFAULT_P_D,
    replicates = N_REPLICATES
  )
}

# Supplemental Fig. S4: no-dropout mutation-rate benchmark.
# Variable: mutation rate mu = 0.05, 0.1, 0.2.
# Both stochastic and heritable dropout rates are set to zero.
if (INCLUDE_FIGS4_NO_DROPOUT) {
  grid_list$figs4_no_dropout <- base_grid(
    pool = "figs4_no_dropout",
    figure = "FigS4",
    panel = "A-C",
    x_variable = "mutation_rate_no_dropout",
    ncells = DEFAULT_NCELLS,
    N_char = DEFAULT_N_CHAR,
    mu = c(0.05, 0.1, 0.2),
    p_ds = NO_DROPOUT_P_DS,
    p_d = NO_DROPOUT_P_D,
    replicates = N_REPLICATES
  )
}

param_grid <- do.call(rbind, grid_list)

# TedSim / expression-generation parameters used by the original generation logic.
param_grid$ngenes <- 500
param_grid$max_walk <- 3
param_grid$p_a <- 0.8
param_grid$n_cif <- 30
param_grid$n_diff <- 20
param_grid$cif_step <- 1
param_grid$ge_prob <- 0.3
param_grid$Sigma <- 0.5
param_grid$N_ms <- 10
param_grid$r_n <- ceiling(param_grid$N_char * 0.7)
param_grid$seed <- GLOBAL_SEED + seq_len(nrow(param_grid)) * 1000L

param_grid$case_id <- mapply(
  make_case_id,
  pool = param_grid$pool,
  ncells = param_grid$ncells,
  N_char = param_grid$N_char,
  mu = param_grid$mu,
  p_ds = param_grid$p_ds,
  p_d = param_grid$p_d,
  replicate = param_grid$replicate
)

param_grid <- param_grid[!duplicated(param_grid$case_id), ]
param_grid$case_index <- seq_len(nrow(param_grid))

planned_file <- file.path(OUT_ROOT, "_manifests", "planned_cases.csv")
safe_write_csv(param_grid, planned_file, row.names = FALSE)

message("Total planned cases: ", nrow(param_grid))
print(table(param_grid$pool))
print(param_grid[, c("case_index", "case_id", "figure", "panel", "x_variable",
                     "ncells", "N_char", "mu", "p_ds", "p_d", "replicate")])
head(param_grid)
```

# One-case generator

```{r one-case-generator}
rename_tree_tips_to_cells <- function(tree, ncells) {
  # TedSim often uses t1, t2, ...; normalize to cell_1, cell_2, ...
  if (length(tree$tip.label) < ncells) {
    stop("Ground-truth tree has fewer tips than ncells.")
  }
  tree$tip.label[seq_len(ncells)] <- paste0("cell_", seq_len(ncells))
  tree
}

generate_one_case <- function(params, out_root = OUT_ROOT) {
  params <- as.list(params)
  case_dir <- file.path(out_root, params$pool, params$case_id)
  case_manifest_file <- file.path(case_dir, "manifest.csv")

  if (dir.exists(case_dir)) {
    if (isTRUE(SKIP_EXISTING_CASES)) {
      message("\n[Skip existing] ", params$case_id)
      message("Existing output: ", normalizePath(case_dir, winslash = "/", mustWork = FALSE))

      if (file.exists(case_manifest_file)) {
        manifest <- read.csv(case_manifest_file, stringsAsFactors = FALSE)
        manifest$generation_status <- "skipped_existing"
        return(manifest)
      }

      return(data.frame(
        case_id = params$case_id,
        case_index = params$case_index,
        pool = params$pool,
        figure = params$figure,
        panel = params$panel,
        x_variable = params$x_variable,
        ncells = params$ncells,
        N_char = params$N_char,
        mu = params$mu,
        p_ds = params$p_ds,
        p_d = params$p_d,
        replicate = params$replicate,
        seed = params$seed,
        case_dir = normalizePath(case_dir, winslash = "/", mustWork = FALSE),
        generation_status = "skipped_existing_no_manifest",
        warning = "Case directory exists but manifest.csv was not found; no files were overwritten.",
        stringsAsFactors = FALSE
      ))
    }

    if (isTRUE(OVERWRITE_CASES)) {
      message("\n[Overwrite existing] ", params$case_id)
      unlink(case_dir, recursive = TRUE, force = TRUE)
    } else {
      stop(
        "Case directory already exists: ", case_dir,
        ". Set SKIP_EXISTING_CASES <- TRUE to skip it, or OVERWRITE_CASES <- TRUE to regenerate it."
      )
    }
  }

  dir.create(case_dir, recursive = TRUE, showWarnings = FALSE)

  message("\n[Generating] ", params$case_id)
  message("Output: ", normalizePath(case_dir, winslash = "/", mustWork = FALSE))

  # Fixed cell-state tree from the original Rmd.
  cell_state_tree_text <- "((((((t1:1)n9:1,(t2:1)n10:1)n8:1)n7:1,(((t3:1)n13:1)n12:1)n11:1)n6:1)n5:1)n4;"
  cell_state_tree <- ape::read.tree(text = cell_state_tree_text)
  state_lineages <- make_state_lineages(cell_state_tree)
  s_depth <- state_depths(cell_state_tree)
  terminal_states <- seq_along(cell_state_tree$tip.label)

  # Generate raw lineage barcodes, cell states, CIFs, and true tree using original TedSim logic.
  set.seed(params$seed)
  returnlist <- SIFGenerate(cell_state_tree, params$n_diff, step = params$cif_step)

  cifs <- simulate_cifs_retry(
    ncells = params$ncells,
    cell_state_tree = cell_state_tree,
    p_a = params$p_a,
    n_CIF = params$n_cif,
    n_diff = params$n_diff,
    step = params$cif_step,
    p_d = params$p_d,
    Sigma = params$Sigma,
    mu = params$mu,
    N_char = params$N_char,
    max_walk_candidates = unique(c(params$max_walk, MAX_WALK_CANDIDATES)),
    SIF_res = returnlist,
    unif_on = FALSE,
    n_retries = SIMULATE_CIFS_RETRIES,
    seed = params$seed
  )

  params$simulate_attempt <- attr(cifs, "simulate_attempt")
  params$max_walk_used <- attr(cifs, "max_walk_used")
  params$seed_used <- attr(cifs, "seed_used")
  params$simulator_backend <- "TedSim_SimulateCIFs"

  barcodes_all <- cifs[[7]]

  # CIF list for leaves. Original code assumes three terminal cell states.
  n_terminal <- length(cell_state_tree$tip.label)
  cif_leaves <- vector("list", n_terminal)
  for (i in seq_len(n_terminal)) {
    cif_leaves[[i]] <- cifs[[1]][[i]][seq_len(params$ncells), , drop = FALSE]
  }

  true_counts_res <- CIF2Truecounts(
    ngenes = params$ngenes,
    ncif = params$n_cif,
    ge_prob = params$ge_prob,
    ncells = params$ncells,
    cif_res = list(cif_leaves, cifs[[2]])
  )

  counts_true <- t(true_counts_res[[1]])
  states_leaves <- cifs[[2]][seq_len(params$ncells), , drop = FALSE]

  cell_ids <- paste0("cell_", states_leaves[, "cellID"])
  rownames(counts_true) <- cell_ids
  colnames(counts_true) <- paste0("Gene", seq_len(ncol(counts_true)))

  counts_expr <- counts_true
  expression_backend <- "TedSim_CIF2Truecounts"

  if (isTRUE(SIMULATE_OBSERVED_COUNTS) && exists("True2ObservedCounts", mode = "function", inherits = TRUE)) {
    obs_try <- tryCatch(
      True2ObservedCounts(counts_true),
      error = function(e) {
        warning("True2ObservedCounts failed: ", conditionMessage(e), ". Using true counts as expression.")
        NULL
      }
    )
    if (!is.null(obs_try)) {
      counts_expr <- obs_try
      expression_backend <- "TedSim_True2ObservedCounts"
    }
  }
  params$expression_backend <- expression_backend

  # Ground-truth tree.
  tree_ct <- cifs[[4]]
  tree_ct <- rename_tree_tips_to_cells(tree_ct, params$ncells)

  # Barcode matrices.
  barcodes_initial <- barcodes_all[seq_len(params$ncells), , drop = FALSE]
  barcode_true <- standardize_barcode_matrix(barcodes_initial, cell_ids = cell_ids, prefix = "S")

  qc <- barcode_qc(barcode_true, params)
  params <- c(params, qc)

  dropout_res <- add_controlled_stochastic_dropout(
    barcode_true,
    p_ds = params$p_ds,
    seed = params$seed + 777L
  )
  barcode_observed <- dropout_res$barcode_observed
  barcode_dropout_type <- dropout_res$dropout_type


  # barcode_imputed <- try_fatescape_imputation(
  #   barcode_observed = barcode_observed,
  #   N_char = params$N_char,
  #   ncells = params$ncells,
  #   r_n = params$r_n
  # )


  # Metadata.
  state_label <- as.integer(states_leaves[, "cluster"])
  state_type <- ifelse(state_label %in% terminal_states, "terminal", "internal_or_root")
  state_depth <- as.integer(s_depth[as.character(state_label)])

  cell_metadata <- data.frame(
    cell_id = cell_ids,
    state_label = state_label,
    state_type = state_type,
    state_depth = state_depth,
    ted_parent_state = as.integer(states_leaves[, "parent"]),
    ted_cluster = as.integer(states_leaves[, "cluster"]),
    ted_depth = as.integer(states_leaves[, "depth"]),
    ted_cellID = as.integer(states_leaves[, "cellID"]),
    stringsAsFactors = FALSE
  )

  represented_states <- sort(unique(state_label))
  params$n_sampled_states <- length(represented_states)
  params$n_sampled_terminal_states <- sum(represented_states %in% terminal_states)
  params$n_sampled_internal_states <- sum(!(represented_states %in% terminal_states))
  params$terminal_state_cell_fraction <- mean(state_type == "terminal")
  params$internal_state_cell_fraction <- mean(state_type == "internal_or_root")
  params$sampled_states <- paste(represented_states, collapse = ";")

  # Canonical standardized outputs only.
  # Method-specific inputs should be produced later by converter scripts from these files.
  expression_file <- file.path(case_dir, "expression.csv")
  true_counts_file <- file.path(case_dir, "true_counts.csv")
  barcode_true_file <- file.path(case_dir, "barcode_true.csv")
  barcode_observed_file <- file.path(case_dir, "barcode_observed.csv")
  # barcode_imputed_file <- file.path(case_dir, "barcode_imputed_fatescape.csv")
  barcode_dropout_type_file <- file.path(case_dir, "barcode_dropout_type.csv")
  metadata_file <- file.path(case_dir, "cell_metadata.csv")
  true_tree_file <- file.path(case_dir, "true_tree.nwk")
  state_tree_file <- file.path(case_dir, "state_tree.nwk")
  state_lineages_file <- file.path(case_dir, "state_lineages.rds")

  safe_write_csv(as.data.frame(counts_expr, check.names = FALSE), expression_file, row.names = TRUE)
  safe_write_csv(as.data.frame(counts_true, check.names = FALSE), true_counts_file, row.names = TRUE)
  safe_write_csv(as.data.frame(barcode_true, check.names = FALSE), barcode_true_file, row.names = TRUE)
  safe_write_csv(as.data.frame(barcode_observed, check.names = FALSE), barcode_observed_file, row.names = TRUE)
  # safe_write_csv(as.data.frame(barcode_imputed, check.names = FALSE), barcode_imputed_file, row.names = TRUE)
  safe_write_csv(as.data.frame(barcode_dropout_type, check.names = FALSE), barcode_dropout_type_file, row.names = TRUE)
  safe_write_csv(cell_metadata, metadata_file, row.names = FALSE)
  safe_write_tree(tree_ct, true_tree_file)
  safe_write_tree(cell_state_tree, state_tree_file)
  safe_save_rds(state_lineages, state_lineages_file)

  # RDS with original TedSim objects.
  safe_save_rds(
    list(
      cifs = cifs,
      returnlist = returnlist,
      true_counts_res = true_counts_res,
      state_lineages = state_lineages,
      params = params
    ),
    file.path(case_dir, "tedsim_objects.rds")
  )

  params_df <- as.data.frame(params, stringsAsFactors = FALSE)
  safe_write_csv(params_df, file.path(case_dir, "params.csv"), row.names = FALSE)

  manifest <- data.frame(
    case_id = params$case_id,
    case_index = params$case_index,
    pool = params$pool,
    figure = params$figure,
    panel = params$panel,
    x_variable = params$x_variable,
    ncells = params$ncells,
    N_char = params$N_char,
    mu = params$mu,
    p_ds = params$p_ds,
    p_d = params$p_d,
    replicate = params$replicate,
    seed = params$seed,
    seed_used = params$seed_used,
    max_walk_used = params$max_walk_used,
    simulate_attempt = params$simulate_attempt,
    simulator_backend = params$simulator_backend,
    expression_backend = params$expression_backend,
    case_dir = normalizePath(case_dir, winslash = "/", mustWork = TRUE),
    expression_file = normalizePath(expression_file, winslash = "/", mustWork = TRUE),
    true_counts_file = normalizePath(true_counts_file, winslash = "/", mustWork = TRUE),
    barcode_true_file = normalizePath(barcode_true_file, winslash = "/", mustWork = TRUE),
    barcode_observed_file = normalizePath(barcode_observed_file, winslash = "/", mustWork = TRUE),
    # barcode_imputed_file = normalizePath(barcode_imputed_file, winslash = "/", mustWork = TRUE),
    barcode_dropout_type_file = normalizePath(barcode_dropout_type_file, winslash = "/", mustWork = TRUE),
    metadata_file = normalizePath(metadata_file, winslash = "/", mustWork = TRUE),
    true_tree_file = normalizePath(true_tree_file, winslash = "/", mustWork = TRUE),
    state_tree_file = normalizePath(state_tree_file, winslash = "/", mustWork = TRUE),
    state_lineages_file = normalizePath(state_lineages_file, winslash = "/", mustWork = TRUE),
    params_file = normalizePath(file.path(case_dir, "params.csv"), winslash = "/", mustWork = TRUE),
    tedsim_objects_file = normalizePath(file.path(case_dir, "tedsim_objects.rds"), winslash = "/", mustWork = TRUE),
    n_mutated_entries_true = params$n_mutated_entries_true,
    n_missing_entries_true = params$n_missing_entries_true,
    mutation_fraction_true = params$mutation_fraction_true,
    missing_fraction_true = params$missing_fraction_true,
    n_sampled_states = params$n_sampled_states,
    n_sampled_terminal_states = params$n_sampled_terminal_states,
    n_sampled_internal_states = params$n_sampled_internal_states,
    terminal_state_cell_fraction = params$terminal_state_cell_fraction,
    internal_state_cell_fraction = params$internal_state_cell_fraction,
    sampled_states = params$sampled_states,
    generation_status = "generated",
    stringsAsFactors = FALSE
  )

  safe_write_csv(manifest, file.path(case_dir, "manifest.csv"), row.names = FALSE)

  message("[Done] ", params$case_id)
  manifest
}
```

# Batch generation

```{r batch-generation}
selected_indices <- parse_case_indices(CASE_INDICES, nrow(param_grid))
selected_indices <- selected_indices[selected_indices >= 1 & selected_indices <= nrow(param_grid)]
selected_cases <- param_grid[selected_indices, , drop = FALSE]

message("Selected cases: ", nrow(selected_cases))
print(selected_cases[, c("case_index", "case_id", "pool", "figure", "panel", "x_variable",
                            "ncells", "N_char", "mu", "p_ds", "p_d", "replicate")])

batch_manifest <- data.frame()

if (isTRUE(RUN_GENERATION)) {
  for (ii in seq_len(nrow(selected_cases))) {
    params <- selected_cases[ii, ]

    one <- tryCatch(
      generate_one_case(params, OUT_ROOT),
      error = function(e) {
        msg <- conditionMessage(e)
        message("[ERROR] ", params$case_id, ": ", msg)
        if (isTRUE(STOP_ON_ERROR)) stop(e)
        data.frame(
          case_id = params$case_id,
          case_index = params$case_index,
          pool = params$pool,
          figure = params$figure,
          panel = params$panel,
          x_variable = params$x_variable,
          ncells = params$ncells,
          N_char = params$N_char,
          mu = params$mu,
          p_ds = params$p_ds,
          p_d = params$p_d,
          replicate = params$replicate,
          seed = params$seed,
          generation_status = "error",
          error = msg,
          stringsAsFactors = FALSE
        )
      }
    )

    batch_manifest <- rbind_fill_base(batch_manifest, one)
  }

  batch_file <- file.path(
    OUT_ROOT, "_manifests",
    paste0("batch_manifest_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".csv")
  )
  safe_write_csv(batch_manifest, batch_file, row.names = FALSE)
  message("Batch manifest written to: ", batch_file)
} else {
  message("RUN_GENERATION is FALSE. Only parameter grid was generated.")
}
```

# Collect all generated manifests

```{r collect-manifests}
manifest_files <- list.files(
  OUT_ROOT,
  pattern = "^manifest\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)

manifest_files <- manifest_files[!grepl("_manifests", manifest_files)]

if (length(manifest_files) == 0) {
  message("No generated case manifests found yet.")
} else {
  all_manifest <- do.call(
    rbind,
    lapply(manifest_files, function(f) {
      read.csv(f, stringsAsFactors = FALSE)
    })
  )

  all_manifest_file <- file.path(OUT_ROOT, "_manifests", "all_cases_manifest.csv")
  safe_write_csv(all_manifest, all_manifest_file, row.names = FALSE)

  message("All-case manifest written to: ", all_manifest_file)
  print(table(all_manifest$pool))

  message("First case directory: ", all_manifest$case_dir[1])
  print(list.files(all_manifest$case_dir[1], recursive = FALSE))
}
```

# How to subset canonical cases for downstream revision experiments

```{r downstream-subsets, eval=FALSE}
manifest <- read.csv(
  file.path(OUT_ROOT, "_manifests", "all_cases_manifest.csv"),
  stringsAsFactors = FALSE
)

# Ablation: use the core 1024-cell cases.
ablation_cases <- subset(
  manifest,
  pool == "core" &
    ncells == 1024 &
    N_char == 16 &
    mu %in% c(0.1, 0.2, 0.3) &
    p_ds %in% c(0.1, 0.2, 0.3)
)

# Parameter sensitivity: use main and stress conditions.
param_cases_main <- subset(
  manifest,
  pool == "core" & ncells == 1024 & N_char == 16 &
    mu == 0.2 & p_ds == 0.2
)

param_cases_stress <- subset(
  manifest,
  pool == "core" & ncells == 1024 & N_char == 16 &
    mu == 0.1 & p_ds == 0.3
)

# Runtime: use n=512/1024/2048 core cases and optional n=256 cases.
runtime_cases <- subset(
  manifest,
  pool %in% c("core", "runtime_optional") &
    N_char == 16 & mu == 0.2 & p_ds == 0.2 &
    replicate <= 3
)

# TideTree / TediTree subset. Method-specific input files should be generated later by converters.
tidetree_cases <- subset(
  manifest,
  pool == "core" & ncells == 512 & N_char == 16 & replicate <= 5
)

# Imputation analysis.
imputation_cases <- subset(
  manifest,
  (pool == "core" & ncells == 1024 & N_char == 16) |
    pool == "imputation_stress"
)
```
